---
title: "Electricity Generation in the Philippines"
author: "Your Name"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(gt)

energy <- read_csv("energy.csv", col_names = c("Source", "Value", "Year", "Units"), skip = 1)

latest_year <- max(energy$Year)

recent <- energy |>
  filter(Year == latest_year) |>
  arrange(desc(Value))

renewables <- c("Geothermal", "Hydropower", "Solar PV", "Wind", "Biofuels", "Waste")

total_by_source <- energy |>
  group_by(Source) |>
  summarise(
    `First Year` = min(Year),
    `Latest Year` = max(Year),
    `Min (GWh)` = min(Value),
    `Max (GWh)` = max(Value),
    `Latest (GWh)` = Value[Year == max(Year)],
    .groups = "drop"
  ) |>
  arrange(desc(`Latest (GWh)`))

renewable_share <- energy |>
  mutate(Type = if_else(Source %in% renewables, "Renewable", "Non-Renewable")) |>
  group_by(Year, Type) |>
  summarise(Value = sum(Value), .groups = "drop") |>
  pivot_wider(names_from = Type, values_from = Value) |>
  mutate(Share = round(Renewable / (Renewable + `Non-Renewable`) * 100, 1))

latest_share <- renewable_share |> filter(Year == latest_year) |> pull(Share)
earliest_share <- renewable_share |> filter(Year == min(Year)) |> pull(Share)
top_source <- recent |> slice_head(n = 1) |> pull(Source)
top_gwh <- recent |> slice_head(n = 1) |> pull(Value)
```

::: {.callout-important}

## Key Findings at a Glance

In `r latest_year`, **`r top_source`** remained the dominant electricity source at `r scales::comma(top_gwh)` GWh. Renewable sources accounted for **`r latest_share`%** of total generation, up from `r earliest_share`% in 2000, reflecting two decades of gradual energy transition.

:::

## 1. Generation by Source in `r latest_year`

A snapshot of the current energy mix reveals the relative contribution of each source. This section shows total generation (GWh) by source for the most recent year in the dataset, ordered from largest to smallest.

```{r}
#| label: tbl-latest
#| tbl-cap: "Electricity Generation by Source"
recent |>
  mutate(Type = if_else(Source %in% renewables, "Renewable", "Non-Renewable")) |>
  select(Source, Type, `GWh` = Value) |>
  gt() |>
  fmt_number(columns = GWh, decimals = 0) |>
  cols_label(Source = "Energy Source", Type = "Category") |>
  tab_header(title = paste("Electricity Generation by Source in", latest_year))
```

Coal alone accounts for the largest share of generation, while geothermal and hydropower are the leading renewable contributors. Solar PV and wind remain small but are growing rapidly from a low base.

## 2. Historical Range by Source (2000–`r latest_year`)

Examining the full time range shows how each source has evolved. This table summarises the minimum, maximum, and latest annual generation for each source, indicating which technologies have expanded or contracted over two decades.

```{r}
#| label: tbl-summary
#| tbl-cap: "Historical Generation Summary by Source"
total_by_source |>
  gt() |>
  fmt_number(columns = c(`Min (GWh)`, `Max (GWh)`, `Latest (GWh)`), decimals = 0) |>
  tab_header(title = "Historical Generation Summary by Source (GWh)")
```

Coal shows the steepest growth, roughly tripling from its 2000 minimum to its recent peak. Solar PV and wind both record near-zero minimums, reflecting their late entry into the grid, and have grown substantially in recent years.

## 3. Renewable Share of Total Generation Over Time

While absolute generation volumes tell one story, the renewable share of the total mix reveals the pace of the energy transition. This chart plots the percentage of generation from renewable sources each year from 2000 to `r latest_year`.

```{r}
#| label: fig-share
#| fig-cap: "Renewable Share of Total Electricity Generation (%)"
ggplot(renewable_share, aes(x = Year, y = Share)) +
  geom_line(color = "#2E8B57", linewidth = 1.1) +
  geom_point(color = "#2E8B57", size = 2.5) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "grey50") +
  annotate("text", x = min(renewable_share$Year) + 0.3, y = 51.5,
           label = "50% threshold", hjust = 0, size = 3.2, color = "grey40") +
  labs(y = "Renewable Share (%)", x = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) +
  scale_y_continuous(limits = c(0, 80), labels = scales::label_percent(scale = 1))
```
